Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-02-16 submission/all_models.tex
some typos/tweaks, a few more to come

We choose ordinary linear regression (OLS) to quantity the relationship between quality metrics (computed with the \emph{WeightWatcher} tool) 
and the reported test error and/or accuracy metrics.
We regress the metrics on the Top1 (and Top5) reported errors (as dependent variables).
These include Top5 errors for the ImageNet-1K model, percent error for the CIFAR-10/100, SVHN, CUB-200-2011 models, and Pixel accuracy
(Pix.Acc.) and Intersection-Over-Union (IOU) for other models.
We regress them individually on each of the norm-based and PL-based metrics, as described in Section~\ref{sxn:cv}.

>>>>

We choose ordinary least squares (OLS) regression to quantify the relationship between quality metrics (computed with the \emph{WeightWatcher} tool) 
and the reported test error and/or accuracy metrics.
We regress the metrics on the Top1 (and Top5) reported errors (as dependent variables).
These include Top5 errors for the ImageNet-1K model, percent error for the CIFAR-10/100, SVHN, CUB-200-2011 models, and Pixel accuracy
(Pix.Acc.) and Intersection-Over-Union (IOU) for other models.
We regress them individually on each of the norm-based and PL-based metrics, as described in Section~\ref{sxn:cv}.

### 2020-02-16 submission/background.tex
last par, whitespace, misc stuff
pass to convert to TR format, e.g, whitespace, etc.
some typos/tweaks, a few more to come

\section{Background and Related Work}
\label{sxn:background}

@@

\paragraph{Statistical mechanics theory for DNNs.}

Statistical mechanics ideas have long had influence on DNN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (those using fitted PL exponents) are based on statistical mechanics~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}, in particular the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.  
We emphasize that the way in which we (and HT-SR Theory) \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We \emph{use} statistical mechanics in a broader sense, drawing upon techniques from quantitative finance and random matrix theory.  % (RMT).
Thus, much more relevant for our metholodogical approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statisical mechanics of heavy tailed and strongly correlated systems.
% Probably best not to cite at this point %\charles{, and current work (in progress)\cite{blog}}

\paragraph{Norm-based capacity control theory.}

>>>>

%KDD% \vspace{-1mm}
\section{Background and Related Work}
\label{sxn:background}
%\vspace{-1mm}

@@

%KDD% \vspace{-1mm}
\paragraph{Statistical mechanics theory for DNNs.}

Statistical mechanics ideas have long had influence on DNN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (those using fitted PL exponents) are based on statistical mechanics~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}, in particular the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.  
We emphasize that the way in which we (and HT-SR Theory) \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We \emph{use} statistical mechanics in a broader sense, drawing upon techniques from quantitative finance and random matrix theory.  % (RMT).
Thus, much more relevant for our metholodogical approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statistical mechanics of heavy tailed and strongly correlated systems.
% Probably best not to cite at this point %\charles{, and current work (in progress)\cite{blog}}

%KDD% \vspace{-1mm}
\paragraph{Norm-based capacity control theory.}

### 2020-02-16 submission/conclusion.tex
pass to convert to TR format, e.g, whitespace, etc.
some typos/tweaks, a few more to come

\vspace{-1mm}
\section{Conclusion}
\label{sxn:conc}
%\vspace{-1mm}

We have developed (based on strong theory) and evaluated (on a large corpus of publicly-available pretrained models from CV and NLP) methods to predict trends in the quality of state-of-the-art neural networks---without access to training or testing data.
Prior to our work, it was not obvious that norm-based metrics would perform well to predict trends in quality \emph{across} models (as they are usually used \emph{within} a given model or parameterized model class, e.g., to bound generalization error or to construct regularizers).
Our results are the first to demonstrate that they can be used for this important practical problem.
That PL-based metrics perform better (than norm-based metrics) should not be surprising---at least to those familiar with the statisical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} (since our use of PL exponents is designed to capture the idea that well-trained models capture correlations over many size scales in the data).
Again, though, our results are the first to demonstrate this.
It is also gratifying that this approach can be used to provide fine-scale insight (such as rationalizing the flow of correlations or the collapse of size scale) throughout a network.

We conclude with a few comments on what a \emph{practical theory} of DNNs should look like.
To do so, we distinguish between two types of theories:
\emph{non-empirical or analogical theories}, in which one creates, often from general principles, a very simple toy model that can be analyzed rigorously, and one then argues that the model is relevant to the system of interest; and 
\emph{semi-empirical theories}, in which there exists a rigorous asymptotic theory, which comes with parameters, for the system of interest, and one then adjusts or fits those parameters to the finite non-asymptotic data.
A drawback of the former approach is that it typically makes very strong assumptions on the data, and the strength of those assumptions can limit the practical applicability of the theory.
Much of the work on the theory of DNNs focuses on the former type of theory.
Our approach focuses on the latter type of theory.
Our results, which are based on our \emph{use} of sophisticated statistical mechanics theory to solve important practical DNN problems, suggests that our approach should be of interest more generally for those interested in developing a practical DNN theory.

>>>>

%KDD% \vspace{-1mm}
\section{Conclusion}
\label{sxn:conc}
%\vspace{-1mm}

We have developed (based on strong theory) and evaluated (on a large corpus of publicly-available pretrained models from CV and NLP) methods to predict trends in the quality of state-of-the-art neural networks---without access to training or testing data.
Prior to our work, it was not obvious that norm-based metrics would perform well to predict trends in quality \emph{across} models (as they are usually used \emph{within} a given model or parameterized model class, e.g., to bound generalization error or to construct regularizers).
Our results are the first to demonstrate that they can be used for this important practical problem.
That PL-based metrics perform better (than norm-based metrics) should not be surprising---at least to those familiar with the statisical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} (since our use of PL exponents is designed to capture the idea that well-trained models capture correlations over many size scales in the data).
Again, though, our results are the first to demonstrate this.
It is also gratifying that our approach can be used to provide fine-scale insight (such as rationalizing the flow of correlations or the collapse of size scale) throughout a network.

We conclude with a few comments on what a \emph{practical theory} of DNNs should look like.
To do so, we distinguish between two types of theories:
\emph{non-empirical or analogical theories}, in which one creates, often from general principles, a very simple toy model that can be analyzed rigorously, and one then argues that the model is relevant to the system of interest; and 
\emph{semi-empirical theories}, in which there exists a rigorous asymptotic theory, which comes with parameters, for the system of interest, and one then adjusts or fits those parameters to the finite non-asymptotic data.
A drawback of the former approach is that it typically makes very strong assumptions on the data, and the strength of those assumptions can limit the practical applicability of the theory.
Nearly all of the work on the theory of DNNs focuses on the former type of theory.
Our approach focuses on the latter type of theory.
Our results, which are based on our \emph{use} of sophisticated statistical mechanics theory to solve important practical DNN problems, suggests that our approach should be of interest more generally for those interested in developing a practical DNN theory.

### 2020-02-16 submission/cv_models.tex
some typos/tweaks, a few more to come

We can learn much more about a pretrained model by going beyond average values of quality metrics to examining quality metrics for each layer weight matrix, $\mathbf{W}$, as a function of depth (or layer id).  % in the network. 
%The most interesting results are seen when we 
For example, we can 
plot (just) the PL exponent, $\alpha$, for each layer, %
%\footnote{That is, here we consider just $\alpha$ for each layer, i.e., not $\hat{\alpha}$ or $\Vert\mathbf{W}\Vert^{\alpha}_{\alpha}$.} 
as a function of depth.
%
See Figure~\ref{fig:3models-alpha-layers}, which plots $\alpha$ for each layer (the first layer corresponds to data, the last layer to labels) for the least accurate (shallowest) and most accurate (deepest) model in each of the VGG (no BN), ResNet, and DenseNet series.
(Again, much more detailed set of plots is available at our repo; but note that the corresponding layer-wise plots for Frobenius and Spectral norms are much less interesting than the results we present here.)

@@

We consider ResNet20, trained on CIFAR10, before and after applying the Group Regularization distillation technique, as implemented in the \texttt{distiller} package~\cite{distiller}.
We analyze the pretrained 4D\_regularized\_5Lremoved baseline and fine-tuned models. 
The reported baseline test accuracies (Top1$=91.45$ and Top5$=99.75$) are better than the reported fine-tuned test accuracies (Top1$=91.02$ and Top5$=99.67$).  Because the baseline accuracy is greater,  the previous results on ResNet (Table~\ref{table:cv-models} and Figure~\ref{fig:cv2-accuracy}) suggests that the baseline Spectral Norms should be \emph{smaller} on average than the fine-tuned ones. \emph{The oppopsite is observed.}
%\michael{Do we have something backwards here: where we observed that the average norm to increase with decreasing test error (as it ``should'' not), whereas the average PL exponent $\alpha$ decreases (as ``expected'' from HT-SR Theory).  }
%In both cases (Frobenius norm results not shown), we observe the opposite.
%%\charles{WHy is this confusing? Call me...}
Figure~\ref{fig:resnet204D5L} presents the Spectral Norm (here denoted $\log\lambda_{max}$) and PL exponent ($\alpha$) for each individual layer weight matrix $\mathbf{W}$.%
\footnote{Here, we only include layer matrices or feature maps with $M\ge50$.}
On the other hand, the $\alpha$ values (in Figure~\ref{fig:resnet204Dalpha}) do not differ systematically between the baseline and fine-tuned models.
%\michael{Some comment about big jump in norm; but fix previous backwards question first.}--SEE BELOW
Also (not shown), the average (unweighted) baseline $\alpha$ is \emph{smaller} than the fine-tuned average (as predicted by HT-SR Theory, the basis of $\hat{\alpha}$).

@@

The pretrained models in the \texttt{distiller} package have passed some quality metric, but they are much less well trodden than any of the 
VGG, ResNet, or DenseNet series.  %% we considered above.
While norms make good regularizers for a single model, there is no reason \emph{a priori} to expect them correlate so well with test accuracies across different models.
We do expect, however, the PL $\alpha$ to do so because it effectively measures the amount of correlation in the model~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
The reason for the anomalous behavior shown in 
Figure~\ref{fig:resnet204D5L}
is that the \texttt{distiller} Group Regularization technique 
%%has the unusual effect of increasing
causes the norms of the $\mathbf{W}$ pre-activation maps for two Conv2D layers to spuriously increase.
This is difficult to diagnose by analyzing training/test curves, but it is easy to diagnose with our approach.

>>>>

We can learn much more about a pretrained model by going beyond average values of quality metrics to examining quality metrics for each layer weight matrix, $\mathbf{W}$, as a function of depth (or layer id).  % in the network. 
%The most interesting results are seen when we 
For example, we can 
plot (just) the PL exponent, $\alpha$, for each layer, %
%\footnote{That is, here we consider just $\alpha$ for each layer, i.e., not $\hat{\alpha}$ or $\Vert\mathbf{W}\Vert^{\alpha}_{\alpha}$.} 
as a function of depth.
%
See Figure~\ref{fig:3models-alpha-layers}, which plots $\alpha$ for each layer (the first layer corresponds to data, the last layer to labels) for the least accurate (shallowest) and most accurate (deepest) model in each of the VGG (no BN), ResNet, and DenseNet series.
(Again, a much more detailed set of plots is available at our repo; but note that the corresponding layer-wise plots for Frobenius and Spectral norms are much less interesting than the results we present here.)

@@

We consider ResNet20, trained on CIFAR10, before and after applying the Group Regularization distillation technique, as implemented in the \texttt{distiller} package~\cite{distiller}.
We analyze the pretrained 4D\_regularized\_5Lremoved baseline and fine-tuned models. 
The reported baseline test accuracies (Top1$=91.45$ and Top5$=99.75$) are better than the reported fine-tuned test accuracies (Top1$=91.02$ and Top5$=99.67$).  Because the baseline accuracy is greater,  the previous results on ResNet (Table~\ref{table:cv-models} and Figure~\ref{fig:cv2-accuracy}) suggest that the baseline Spectral Norms should be \emph{smaller} on average than the fine-tuned ones. \emph{The oppopsite is observed.}
%\michael{Do we have something backwards here: where we observed that the average norm to increase with decreasing test error (as it ``should'' not), whereas the average PL exponent $\alpha$ decreases (as ``expected'' from HT-SR Theory).  }
%In both cases (Frobenius norm results not shown), we observe the opposite.
%%\charles{WHy is this confusing? Call me...}
Figure~\ref{fig:resnet204D5L} presents the Spectral Norm (here denoted $\log\lambda_{max}$) and PL exponent ($\alpha$) for each individual layer weight matrix $\mathbf{W}$.%
\footnote{Here, we only include layer matrices or feature maps with $M\ge50$.}
On the other hand, the $\alpha$ values (in Figure~\ref{fig:resnet204Dalpha}) do not differ systematically between the baseline and fine-tuned models.
%\michael{Some comment about big jump in norm; but fix previous backwards question first.}--SEE BELOW
Also (not shown), the average (unweighted) baseline $\alpha$ is \emph{smaller} than the fine-tuned average (as predicted by HT-SR Theory, the basis of $\hat{\alpha}$).

@@

The pretrained models in the \texttt{distiller} package have passed some quality metric, but they are much less well trodden than any of the 
VGG, ResNet, or DenseNet series.  %% we considered above.
While norms make good regularizers for a single model, there is no reason \emph{a priori} to expect them correlate so well with test accuracies across different models.
We do expect, however, the PL $\alpha$ to do so because it effectively measures the amount of correlation in the model~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
The reason for the anomalous behavior shown in 
Figure~\ref{fig:resnet204D5L}
is that the \texttt{distiller} Group Regularization technique 
%%has the unusual effect of increasing
causes the norms of the $\mathbf{W}$ pre-activation maps for two Conv2D layers to increase spuriously.
This is difficult to diagnose by analyzing training/test curves, but it is easy to diagnose with our~approach.

### 2020-02-16 submission/introduction.tex
pass to convert to TR format, e.g, whitespace, etc.
included new repo ref
some typos/tweaks, a few more to come

\section{Introduction}
\label{sxn:intro}

@@

\paragraph{The WeightWatcher Tool.}

All of our computations were performed with the publicly-available \emph{WeightWatcher} tool (version 0.2.7)~\cite{weightwatcher_package}.
To be fully reproducible, we only examine publically-available, pretrained models, and we also provide all Jupyter and Google Colab notebooks used in an accompanying github repository~\cite{kdd20_sub_repo}.
See Appendix~\ref{sxn:appendix} for details on how to reproduce all results.

\paragraph{Organization of this paper.}

>>>>

%KDD% \vspace{-1mm}
\section{Introduction}
\label{sxn:intro}
%\vspace{-1mm}

@@

%KDD% \vspace{-1mm}
\paragraph{The WeightWatcher Tool.}

All of our computations were performed with the publicly-available \emph{WeightWatcher} tool (version 0.2.7)~\cite{weightwatcher_package}.
To be fully reproducible, we only examine publicly-available, pretrained models, and we also provide all Jupyter and Google Colab notebooks used in an accompanying github repository~\cite{kdd20_sub_repo}.
See Appendix~\ref{sxn:appendix} for details on how to reproduce all results.

%KDD% \vspace{-1mm}
\paragraph{Organization of this paper.}

### 2020-02-16 submission/methods.tex
some typos/tweaks, a few more to come

Assume we are given several pretrained DNNs, e.g., as part of an architecture series.
The models have been trained and evaluated on labeled data $\{d_{i},y_{i}\}\in\mathcal{D}$, using standard techniques.  
The pretained pytroch model files are publicly-available, and the test accuracies have been reported online.  
In this study, we do not have access to this data, and we have not trained any of the models ourselves,
nor have we re-evaluated the test accuracies.
We expect that most well-trained, production-quality models will employ one or more forms of on regularization, such as Batch Normalization (BN), Dropout, etc., and many will also contain additional structure such as Skip Connections, etc. 
Here, we will ignore these details, and will focus only on the pretrained layer weight matrices $\mathbf{W}_{l}$.

@@

The best performing empirical quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$ and/or, equivalently, it's \emph{Empirical Correlation Matrix}: $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$ .%

@@

\begin{itemize}
\item 
Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum_{i=1}^{M} \lambda_{i}$
\item 
Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$
\item 
Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$
\item 
$\alpha$-Norm (or $\alpha$-Shatten Norm):%
\footnote{Notice $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}$. We use $\mathbf{X}$ to emphasize that $\alpha$ depends on the ESD of $\mathbf{X}$.}
 $\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum_{i=1}^{M}\lambda_{i}^{\alpha}$
\end{itemize}
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, and $\lambda_{max}$ is the maximum eigenvalue.
Recall that the eigenvalues are square of the singular values $\sigma_{i}$ of $\mathbf{W}$: $\lambda_{i}=\sigma^{2}_{i}$.
Also, note that we do \emph{not} normalize $\mathbf{X}$ by $1/N$; see Appendix~\ref{sxn:appendix} for a discussion of this issue.

@@

For norm-based metrics, we use the average of the log norm, and to the appropriate power.
Informallly, this amounts to assuming that the layer weight matrices are statistically independent, in which case we can estimate the model complexity $\mathcal{C}$, or test accuracy, with a standard Product Norm (which resembles a data dependent VC complexity),
\begin{equation}
\mathcal{C}\sim\Vert\mathbf{W}_{1}\Vert\times\Vert\mathbf{W}_{2}\Vert \times \cdots \times \Vert\mathbf{W}_{L}\Vert ,
\end{equation}
where $\Vert\cdot\Vert$ is a matrix norm.   
The log complexity,
\begin{equation}
\label{eqn:eqn:sum_log_norm}
\log\mathcal{C} \sim \log\Vert\mathbf{W}_{1}\Vert+\log\Vert\mathbf{W}_{2}\Vert + \cdots + \log\Vert\mathbf{W}_{L}\Vert  ,
\end{equation}
 takes the form of an average Log Norm.
For the \emph{Frobenius Norm metric} and \emph{Spectral Norm metric}, we can use Eqn.~(\ref{eqn:eqn:sum_log_norm}) directly.% 
\footnote{When taking $\log\Vert\mathbf{W}_{l}\Vert_{F}^{2}$, the $2$ comes down and out of the sum, and thus ignoring it only changes the metric by a constant factor.}

>>>>

Assume we are given several pretrained DNNs, e.g., as part of an architecture series.
The models have been trained and evaluated on labeled data $\{d_{i},y_{i}\}\in\mathcal{D}$, using standard techniques.  
The pretained pytroch model files are publicly-available, and the test accuracies have been reported online.  
In this study, we do not have access to this data, and we have not trained any of the models ourselves,
nor have we re-evaluated the test accuracies.
We expect that most well-trained, production-quality models will employ one or more forms of regularization, such as Batch Normalization (BN), Dropout, etc., and many will also contain additional structure such as Skip Connections, etc. 
Here, we will ignore these details, and will focus only on the pretrained layer weight matrices $\mathbf{W}_{l}$.

@@

The best performing empirical quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's \emph{Empirical Correlation Matrix}: $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.%

@@

\begin{itemize}
\item 
Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum_{i=1}^{M} \lambda_{i}$
\item 
Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$
\item 
Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$
\item 
$\alpha$-Norm (or $\alpha$-Shatten Norm):%
\footnote{Notice $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}$. We use $\mathbf{X}$ to emphasize that $\alpha$ depends on the ESD of $\mathbf{X}$.}
 $\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum_{i=1}^{M}\lambda_{i}^{\alpha}$
\end{itemize}
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, and $\lambda_{max}$ is the maximum eigenvalue.
Recall that the eigenvalues are squares of the singular values $\sigma_{i}$ of $\mathbf{W}$: $\lambda_{i}=\sigma^{2}_{i}$.
Also, note that we do \emph{not} normalize $\mathbf{X}$ by $1/N$; see Appendix~\ref{sxn:appendix} for a discussion of this issue.

@@

For norm-based metrics, we use the average of the log norm, and to the appropriate power.
Informally, this amounts to assuming that the layer weight matrices are statistically independent, in which case we can estimate the model complexity $\mathcal{C}$, or test accuracy, with a standard Product Norm (which resembles a data dependent VC complexity),
\begin{equation}
\mathcal{C}\sim\Vert\mathbf{W}_{1}\Vert\times\Vert\mathbf{W}_{2}\Vert \times \cdots \times \Vert\mathbf{W}_{L}\Vert ,
\end{equation}
where $\Vert\cdot\Vert$ is a matrix norm.   
The log complexity,
\begin{equation}
\label{eqn:eqn:sum_log_norm}
\log\mathcal{C} \sim \log\Vert\mathbf{W}_{1}\Vert+\log\Vert\mathbf{W}_{2}\Vert + \cdots + \log\Vert\mathbf{W}_{L}\Vert  ,
\end{equation}
 takes the form of an average Log Norm.
For the \emph{Frobenius Norm metric} and \emph{Spectral Norm metric}, we can use Eqn.~(\ref{eqn:eqn:sum_log_norm}) directly.% 
\footnote{When taking $\log\Vert\mathbf{W}_{l}\Vert_{F}^{2}$, the $2$ comes down and out of the sum, and thus ignoring it only changes the metric by a constant factor.}
